Synthesizing Random Generators via Higher-Order Representations

### 2019-01-16 paper/sections/hrep.tex
Sync

What to say here:

\begin{itemize}
\item Haskell ADTs
\item Type driven generation
\item Limitations
\item Key idea: Higher order representations
\item Canonical representation of the data type
\item Pattern matchings representation
\item Abstract interface representation
\item Composing representations
\end{itemize}

@@

>>>>

In this work, we ...

\begin{code}
data HRep_Val  a = Mk_Val Int
data HRep_Add  a = Mk_Add a a
data HRep_Mul  a = Mk_Mul a a
\end{code}

For this purpose, we employ Haskell's type classes mechanism \tocite{}.
%
We define an evaluation type class $(\Downarrow)$ which specifies how to perform
a single transformation |step| from the higher-order representation back into a
concrete value of the target:

\begin{code}
class (rep down target) where
  step :: rep target -> target
\end{code}

Then, we can define the following overloaded instances of the |step| operation
for the canonical representations of data constructors simply by mapping each
lifted constructor back into its corresponding one.

\begin{code}
instance (HRep_Val down Exp) where
  step (Mk_Val n) = Val n

instance (HRep_Add down Exp) where
  step (Mk_Add x y) = Add x y

instance (HRep_Mul down Exp) where
  step (Mk_Mul x y) = Mul x y
\end{code}

With these individual representations, we can define a type combinator
$(\oplus)$ to compose two representations into a single one:

\begin{code}
data (f oplus g) a = L (f a) | R (g a)
\end{code}

Then, a composite representation can be transformed back into the concrete
target type by pattern matching on the data type variant and applying the |step|
tranformation to the inner representation:

\begin{code}
instance (f oplus g down Exp) where
  step (L f) = step f
  step (R g) = step g
\end{code}

Furthermore, we can define a type combinator ($\otimes$) to tag every
representation with an explicit generation frequency:

\begin{code}
data (f otimes n) a = Freq (f a)
\end{code}

This combinator is evaluated simply by piping the result from the inner
representation.
%
It does not change the evaluation semantics, as it is only considered at
generation time:

\begin{code}
instance (f otimes n down Exp) where
  step (Freq f) = step f
\end{code}

\begin{code}
type HRep_Exp  = HRep_Val oplus'' HRep_Add oplus'' HRep_Mul
\end{code}

\begin{code}
data (f otimes n) a = TagFreq (f a)

instance (f otimes n down Exp) where
  step (TagFreq f) = step f
\end{code}

\begin{code}
data HRep_foo_1  a = Mk_foo_1 a a
data HRep_foo_2  a = Mk_foo_2 Int a
\end{code}

\begin{code}
instance (HRep_foo_1 down Exp) where
  step (Mk_foo_1 x y)
    = Add (Add x (Val 50)) (Add (Val 25) y)

instance (HRep_foo_2 down Exp) where
  step (Mk_foo_2 x y)
    = Mul (Val 50) (Mul (Val x) y)
\end{code}

\begin{code}
data HRep_ten       a = Mk_ten
data HRep_square    a = Mk_square   a
data HRep_minus     a = Mk_minus    a a
\end{code}

\begin{code}
instance (HRep_ten down Exp) where
  step Mk_ten = ten

instance (HRep_square down Exp) where
  step (Mk_square x) = square x

instance (HRep_minus down Exp) where
  step (Mk_minus x y) = minus x y
\end{code}

\begin{code}
type HRep_foo  =       HRep_foo_1
               oplus'  HRep_foo_2
\end{code}

\begin{code}
type HRep_M  =       HRep_ten
             oplus'  HRep_square
             oplus'  HRep_minus
\end{code}

\begin{code}
type Spec_Exp  =       HRep_Exp  otimes 4
               oplus'  HRep_foo  otimes 2
               oplus'  HRep_M
\end{code}

@@

% \begin{code}
% data (Term f) a = TagTerm (f a)

% instance (Term f down Exp) where
%   step (TagTerm f) = step f
% \end{code}

### 2019-01-16 paper/sections/randomtesting.tex
Sync

In this work we identify two sources of complications that appear when using the
aforementioned automatic derivation tools to obtain random data generators:

\begin{enumerate}
\item The target code behaves differently over inputs matching very specific
  patterns.
\item The target code behaves differently according to some dependency over
  subexpressions of its inputs.
\end{enumerate}

To exemplify the first point, suppose we want to use randomly generated |Exp|s
to test a property comprising the following function:

\begin{code}
foo :: Exp -> Exp
foo (Add (Add x (Val 50)) (Add (Val 25) y))
  = error "pattern #1"
foo (Mul (Mul x (Val 50)) (Mul (Val 25) y))
  = error "pattern #2"
foo x = x
\end{code}

This function behaves mostly like an identity function, except for two very
specific patterns of input values defined using three nested levels of pattern
matching.
%
This is, |foo| does not only matches against the root data constructor, but also
against the data constructors of its subexpressions and sub-subexpressions.

If we consider that \quickcheck's default implementation for random |Int|s pick
a number in the interval $[-n, n]$ with uniform distribution, we need to be able
to identify which one is a sensible generation size to set based on wether our
code branches on numbers or not, otherwise our random generator will not be able
to generate random values to match against such numbers, leaving fragments of
code completely untested.
%
Suppose that we pick the a generation size $50$, i.e. the minimum size that is
big enough to produce an |Int| number equal to $50$.
%
Under this considerations, the probability of generating a value matching the
first clause of |foo| (and hence triggering the error) are as follows:

\begin{align*}
  P(match(foo_1))
  \!&&=\; &P(Add) \\
  \!&&*\; &P(Add) * 1 * (P(Val) * (1/100)) \\
  \!&&*\; &P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

\todo[inline]{AM: COMPLETE}

Despite the fact that the previous example might look artificial, branching
against specific patterns of the input data is a common task.
%
For instance, the balancing function of a Red-Black tree need to consider a
quite specific combination of color, left- and right subtrees and sub-subtrees
in order to preserve the height invariant.
%
Moreover, Common Subexpression Elimination (CSE) is a compiler optimization that
needs to consider very specific sequences of instructions that may be regrouped
in a computationally more efficient way.

We believe that if a certain pattern of values is relevant enough to appear in
the codebase being tested, a practical random testing methodology should be able
to produce values satisfying such pattern in a substantial proportion.

\begin{code}
fortyTwo :: Exp
fortyTwo = Val 42

twice :: Exp -> Exp
twice x = Add x x

square :: Exp -> Exp
square x = Mul x x

neg :: Exp -> Exp
neg x = Mul x (Val (-1))

minus :: Exp -> Exp -> Exp
minus x y = Add x (neg y)
\end{code}

% \begin{code}
% data IExp  =   VAL Int
%            pp  ADD IExp IExp
%            pp  IF BExp IExp IExp

% data BExp  =   TRUE | FALSE
%            pp  LE IExp IExp
%            pp  AND BExp BExp
% \end{code}

% \begin{code}
% foo :: IExp -> IExp
% foo (ADD (IF TRUE (VAL 42) x) y)     = ...
% foo (IF (LE (VAL 24) x) (VAL y)) z)  = ...
% foo x                                = ...

% bar :: BExp -> BExp
% bar (LE (ADD (VAL 12) x) (VAL 34))   = ...
% bar (LE (VAL 78) (IF FALSE x y)) z)  = ...
% bar x                                = ...
% \end{code}

% \begin{code}
% twice :: IExp -> IExp
% twice e = ADD e e

% equals :: IExp -> IExp -> BExp
% equals x y = AND (LE x y) (LE y x)

% and3 :: BExp -> BExp -> BExp -> BExp
% and3 x y z = AND x (AND y z)
% \end{code}

>>>>

In this work we identify two sources of additional structure information which
are not considered by the aforementioned automatic derivation tools to obtain
random data generators:

\begin{enumerate}
\item The target code behaves differently over inputs matching very specific
  patterns.
\item The target code yields some of the responsability of preserving its
  invariants to the functions of its abstract interface.
\end{enumerate}

This information can be reified in compile time and used to synthesize richer
random generators automatically. We proceed to exemplify the previous points in
detail.

\subsection{\textbf{Presence of Complex Pattern Matchings}}

To exemplify the firts problematic scenarion that may arise, suppose we want to
use randomly generated |Exp|s to test a property comprising the following
function:

\begin{code}
foo :: Exp -> Exp
foo (Add (Add x (Val 50)) (Add (Val 25) y))
  = error "pattern #1"
foo (Mul (Val 50) (Mul (Val x) y))
  = error "pattern #2"
foo x = x
\end{code}

This function behaves very much like an identity function, with the exception
that it fails for two very specific patterns of input values defined using
nested pattern matching.
%
This is, |foo| does not only matches against the root data constructor, but also
against the data constructors of its subexpressions and sub-subexpressions.

\quickcheck's default implementation for random |Int|s pick a number in the
interval $[-n, n]$ with uniform distribution.
%
Hence, in principle we need to recognize what is a suitable generation size,
which should be big enough for our generator be able to generate the numbers we
pattern match against to, otherwise we will not be able to generate any value to
match against |foo| clauses, leaving fragments of code completely untested.
%
Suppose then that we pick the a generation size $50$, i.e. the minimum size that
is big enough to produce an |Int| number equal to $50$.
%
Under this consideration, the probability of generating a value matching the
first clause of |foo| (and hence triggering the first error) results as follows:

\begin{align*}
  P(match(foo\#1))
  \!&&=\; &P(Add) \\
  \!&&*\; &P(Add) * 1 * (P(Val) * (1/100)) \\
  \!&&*\; &P(Add) * (P(Val) * (1/100)) * 1
\end{align*}

If we use \megadeth to obtain a random generator for |Exp|, we obtain a uniform
generation probability distribution over constructors, i.e., $P(|Val|) =
P(|Add|) = P(|Mul|) = 1/3$.
%
In this setting, $P(match(foo\#1))$ results $1/2430000$, meaning that, in
average, we will need to generate over than two million test cases in order to
be able to test the first clause of |foo| only once.
%
This situation can be somewhat improved if we decide to use \dragen to obtain a
random generator.
%
By optimizing the generation probabilities in order to benefit the generation of
|Add| constructor in a ratio of $20:1$ with respect to the rest of them, we
obtained a distribution of values where $P(match(foo\#1)) \sim 1/300000$.
%
Although this certainly improves the probability of generating a matching value,
this probability is not substantial enough to become practical.
%
Additionally, by favoring the generation probabilities towards the |Add|
constructor, we found that the probability of generating a value matching the
second clause of |foo| (which does not matches against it) also gets diminished
into an impractival value.

Despite the fact that the previous example might look artificial, branching
against specific patterns of the input data is a common task.
%
For instance, the balancing function of a Red-Black tree need to consider a
quite specific combination of color, left and right subtrees and sub-subtrees
in order to preserve the height invariant.
%
Moreover, Common Subexpression Elimination (CSE) is a compiler optimization that
needs to consider very specific sequences of instructions that may be regrouped
in a computationally more efficient way, to cite a few.

\subsection{\textbf{Data Invariants Encoded on Abstract Interfaces}}

A common choice when implementing a data structure is to transfer the
responsability of preserving its invariants to the functions that manipulates
its values.
%
Consider for instance the following possible implementation of a basic |HTML|
manipulation library:

\begin{code}
data Html = Html String

head :: Html -> Html
head (Html inner)
  = Html ("<head>" ++ inner ++ "</head>")

body :: Html -> Html
body (Html inner)
  = Html ("<body>" ++ t ++ "</body>")

br :: Html
br = Html "<br />"

concat :: Html -> Html -> Html
concat (Html x) (Html y) = Html (x ++ y)
\end{code}

In the previous definition, the |Html| data type is defined using a single data
constructor that contains the textual representation of the Html code it
represents in plain text.
%
Later, the combinator functions defined over |HTML| are the ones in charge of
transforming this plain text representation with the invariant that, given valid
|Html| parameters, they always return a valid |Html| value.
%
This is a very commmon programming pattern that can be found in a variety of
Haskell libraries \tocite{}.

If we use a standard type-driven approach to derive a random generator for
|Html|, the only ``structural'' information that we can use in the derivation
process is that |Html| values are composed of strings.
%
As a consequence, we essentially end up generating random strings, which rarely
represents a valid (or at least well structured) Html value.
%
Additionally, if our test suite contains properties constrained by certain
preconditions, this lack of domain knowleadge may lead in an impractically high
discard ratio of randomly generated test.
%
For instance, suposse we write a property to verify that |head| preserves the
invariant previously mentioned:

\begin{code}
prop_head_inv :: Html -> Bool
prop_head_inv x = valid x ===> valid (head x)
\end{code}

In the previous definition we state that |head| always returns a valid |Html|
value (|valid (head x)|), provided that we are given a valid |Html| as input
(|valid x|).
%
Then, while testing this property we rarely satisfy its precondition, in which
case \quickcheck discards the whole test, retrying with a diferent random input,
degradating this way the testing performance.

We believe that, if certain patterns of values are relevant enough to appear in
the codebase being tested, a practical random testing methodology should be able
to produce values satisfying such patterns in a substantial proportion.
%
The next section introduces a reprentation model that let us encode the
structure information presented here into our automatically derived random
generators in a modular and flexible way.

% Consider the following Haskell module |M| which defines combinators for the type
% |Exp|:

% \begin{code}
% module M where

% ten :: Exp
% ten = Val 10

% square :: Exp -> Exp
% square x = Mul x x

% minus :: Exp -> Exp -> Exp
% minus x y = Add x (Mul y (Val (-1)))
% \end{code}

### 2019-01-16 paper/tikz/hrep.tex
Sync

%Up Arrow [id:dp3837429966878143] 
\draw  [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][fill={rgb, 255:red, 248; green, 231; blue, 28 }  ,fill opacity=1 ][line width=1.5]  (360,160) -- (370,150) -- (380,160) -- (380,160) -- (380,160) -- (360,160) -- (360,160) -- cycle ;
%Up Arrow [id:dp0013771056350821986] 
\draw  [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][fill={rgb, 255:red, 248; green, 231; blue, 28 }  ,fill opacity=1 ][line width=1.5]  (260,160) -- (270,150) -- (280,160) -- (280,160) -- (280,160) -- (260,160) -- (260,160) -- cycle ;
%Shape: Rectangle [id:dp7366907361635939] 
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (240.43,50) -- (300,50) -- (300,80) -- (240.43,80) -- cycle ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (210,65) .. controls (210,59.48) and (214.48,55) .. (220,55) .. controls (225.52,55) and (230,59.48) .. (230,65) .. controls (230,70.52) and (225.52,75) .. (220,75) .. controls (214.48,75) and (210,70.52) .. (210,65) -- cycle ; \draw  [line width=1.5]  (210,65) -- (230,65) ; \draw  [line width=1.5]  (220,55) -- (220,75) ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (310,65) .. controls (310,59.48) and (314.48,55) .. (320,55) .. controls (325.52,55) and (330,59.48) .. (330,65) .. controls (330,70.52) and (325.52,75) .. (320,75) .. controls (314.48,75) and (310,70.52) .. (310,65) -- cycle ; \draw  [line width=1.5]  (310,65) -- (330,65) ; \draw  [line width=1.5]  (320,55) -- (320,75) ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (152.93,28.79) .. controls (156.83,24.88) and (163.17,24.88) .. (167.07,28.79) .. controls (170.98,32.69) and (170.98,39.02) .. (167.07,42.93) .. controls (163.17,46.83) and (156.83,46.83) .. (152.93,42.93) .. controls (149.02,39.02) and (149.02,32.69) .. (152.93,28.79) -- cycle ; \draw  [line width=1.5]  (152.93,28.79) -- (167.07,42.93) ; \draw  [line width=1.5]  (167.07,28.79) -- (152.93,42.93) ;
%Snip Same Side Corner Rect [id:dp4221187902899477] 
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (240.1,30.5) -- (250.6,20) -- (289.5,20) -- (300,30.5) -- (300,50) -- (300,50) -- (240.1,50) -- (240.1,50) -- cycle ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (252.93,28.79) .. controls (256.83,24.88) and (263.17,24.88) .. (267.07,28.79) .. controls (270.98,32.69) and (270.98,39.02) .. (267.07,42.93) .. controls (263.17,46.83) and (256.83,46.83) .. (252.93,42.93) .. controls (249.02,39.02) and (249.02,32.69) .. (252.93,28.79) -- cycle ; \draw  [line width=1.5]  (252.93,28.79) -- (267.07,42.93) ; \draw  [line width=1.5]  (267.07,28.79) -- (252.93,42.93) ;
%Shape: Rectangle [id:dp1692119124811613] 
\draw  [line width=1.5]  (130,170) -- (410,170) -- (410,200) -- (130,200) -- cycle ;

% Text Node
\draw (270,185) node [scale=1.2]  {$T$};
% Text Node
\draw (270,125) node [scale=1]  {$\text{Patterns}^{T}_{foo}$};
% Text Node
\draw (370,125) node [scale=1]  {$\text{Interface}^{T}_{M}$};
% Text Node
\draw (170.05,65) node [scale=0.9]  {$\text{HRep}^{T}_{T}$};
% Text Node
\draw (370,65) node [scale=0.9]  {$\text{HRep}^{T}_{M}$};
% Text Node
\draw (114.5,125) node [scale=1.2]  {$S_{T}\Downarrow T$};
% Text Node
\draw (392,25) node [scale=1.2]  {$S_{T}$};
% Text Node
\draw (270.22,65) node [scale=0.9]  {$\text{HRep}^{T}_{foo}$};
% Text Node
\draw (183.05,35) node [scale=1.44]  {$\alpha $};
% Text Node
\draw (284.05,35) node [scale=1.44]  {$\beta $};

>>>>

%Up Arrow [id:dp0013771056350821986]
\draw  [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][fill={rgb, 255:red, 248; green, 231; blue, 28 }  ,fill opacity=1 ][line width=1.5]  (260,149.5) -- (270,140) -- (280,149.5) -- (273.5,149.5) -- (273.5,170) -- (266.5,170) -- (266.5,149.5) -- cycle ;
%Shape: Rectangle [id:dp7366907361635939]
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (240.43,50) -- (300,50) -- (300,80) -- (240.43,80) -- cycle ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (210,65) .. controls (210,59.48) and (214.48,55) .. (220,55) .. controls (225.52,55) and (230,59.48) .. (230,65) .. controls (230,70.52) and (225.52,75) .. (220,75) .. controls (214.48,75) and (210,70.52) .. (210,65) -- cycle ; \draw  [line width=1.5]  (210,65) -- (230,65) ; \draw  [line width=1.5]  (220,55) -- (220,75) ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (310,65) .. controls (310,59.48) and (314.48,55) .. (320,55) .. controls (325.52,55) and (330,59.48) .. (330,65) .. controls (330,70.52) and (325.52,75) .. (320,75) .. controls (314.48,75) and (310,70.52) .. (310,65) -- cycle ; \draw  [line width=1.5]  (310,65) -- (330,65) ; \draw  [line width=1.5]  (320,55) -- (320,75) ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (152.93,28.79) .. controls (156.83,24.88) and (163.17,24.88) .. (167.07,28.79) .. controls (170.98,32.69) and (170.98,39.02) .. (167.07,42.93) .. controls (163.17,46.83) and (156.83,46.83) .. (152.93,42.93) .. controls (149.02,39.02) and (149.02,32.69) .. (152.93,28.79) -- cycle ; \draw  [line width=1.5]  (152.93,28.79) -- (167.07,42.93) ; \draw  [line width=1.5]  (167.07,28.79) -- (152.93,42.93) ;
%Snip Same Side Corner Rect [id:dp4221187902899477]
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (240.1,30.5) -- (250.6,20) -- (289.5,20) -- (300,30.5) -- (300,50) -- (300,50) -- (240.1,50) -- (240.1,50) -- cycle ;
\draw  [fill={rgb, 255:red, 255; green, 255; blue, 255 }  ,fill opacity=1 ][line width=1.5]  (252.93,28.79) .. controls (256.83,24.88) and (263.17,24.88) .. (267.07,28.79) .. controls (270.98,32.69) and (270.98,39.02) .. (267.07,42.93) .. controls (263.17,46.83) and (256.83,46.83) .. (252.93,42.93) .. controls (249.02,39.02) and (249.02,32.69) .. (252.93,28.79) -- cycle ; \draw  [line width=1.5]  (252.93,28.79) -- (267.07,42.93) ; \draw  [line width=1.5]  (267.07,28.79) -- (252.93,42.93) ;
%Shape: Rectangle [id:dp1692119124811613]
\draw  [line width=1.5]  (130,170) -- (410,170) -- (410,200) -- (130,200) -- cycle ;
%Up Arrow [id:dp4999372801455]
\draw  [color={rgb, 255:red, 0; green, 0; blue, 0 }  ,draw opacity=1 ][fill={rgb, 255:red, 248; green, 231; blue, 28 }  ,fill opacity=1 ][line width=1.5]  (360,149.5) -- (370,140) -- (380,149.5) -- (373.5,149.5) -- (373.5,170) -- (366.5,170) -- (366.5,149.5) -- cycle ;

% Text Node
\draw (270,185) node [scale=1.2]  {$Exp$};
% Text Node
\draw (270,125) node [scale=1]  {$\text{Patterns}_{foo}$};
% Text Node
\draw (370,125) node [scale=1]  {$\text{Interface}_{M}$};
% Text Node
\draw (170.05,65) node [scale=1]  {$\text{HRep}_{Exp}$};
% Text Node
\draw (370,65) node [scale=1]  {$\text{HRep}_{M}$};
% % Text Node
% \draw (77.5,125) node [scale=1.2]  {$Spec_{Exp}\Downarrow \ Exp$};
% Text Node
\draw (377.5,25) node [scale=1.2]  {$Spec_{Exp}$};
% Text Node
\draw (270.22,65) node [scale=1]  {$\text{HRep}_{foo}$};
% Text Node
\draw (181.5,34) node [scale=1.44]  {$4$};
% Text Node
\draw (284.05,34) node [scale=1.44]  {$2$};
